---
title: "DECENT DE analysis"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(openxlsx)
library(scater)
library(scran)
library(ReactomePA)
library(ComplexHeatmap)

# Path to project folder
wd <- here::here()
source(file.path(wd, "scRNAseq/analysis/scripts/colors.R"))
source(file.path(wd, "scRNAseq/analysis/scripts/setup.R"))

# Parallelization
library(BiocParallel)
bpp <- MulticoreParam()
```

```{r}
iteration_input = "6_MCL_CNV"
sce.dir <- file.path(wd, "scRNAseq/analysis/outputs", iteration_input)
sce.MCL <- readRDS(file = file.path(sce.dir, "sce.rds"))
sce.MCL
```

```{r, fig.width=12, fig.height=6}
sce.MCL.plot <- sce.MCL
sce.MCL.plot$Sample <- sce.MCL.plot$sample_norep_plot
sce.MCL.plot$Sample <- factor(sce.MCL.plot$Sample, 
                              levels = levels(sce.MCL.plot$sample_norep) %>% str_replace_all(., "_", " "))

plotReducedDim(sce.MCL.plot,
               dimred = "TSNE_corrected_SamplePhase",
               color_by = "patient",
               other_fields = "timepoint") +
  facet_wrap( ~ timepoint) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  xlim(-35, 40) + ylim(-40, 35) +
  cowplot::theme_half_open(font_size = 14) +
  theme(strip.text = element_text(face = "bold")) +
  cowplot::panel_border(color = "grey50") +
  labs(x = "TSNE 1", y = "TSNE 2", color = "Patient")
```

# Pseudobulk DE between timepoints

## Output of `DECENT` package

`DECENT` DE analysis outputs a table of genes compared between conditions with log fold changes, p values and stat value. Here that table is annotated and then plotted.

```{r message=FALSE}
DE.output.path <- file.path(wd, "scRNAseq/analysis/outputs/DE_DECENT_outs")

timepoint.de.list <- list()
samples <- patients[1:4]
for(samplename in samples){
  filename <- paste0("DE_timepoint_", samplename, ".csv")
  timepoint.de.list[[samplename]] <- read_csv(file.path(DE.output.path, filename))
}
```

```{r}
timepoint.de.list %>% 
  bind_rows(.id = "patient") %>% 
  write_csv(
    file = file.path(wd, "scRNAseq/analysis/results", "DE_timepoint_all.csv")
  )
```


```{r message=FALSE}
sig.pvalue = 0.01
sig.logfc = 2

timepoint.de.list <- lapply(
    timepoint.de.list,
    mutate,
    change = case_when(abs(logfc) > sig.logfc & pvalue < sig.pvalue & logfc < 0 ~ "down",
                       abs(logfc) > sig.logfc & pvalue < sig.pvalue & logfc > 0 ~ "up",
                       TRUE ~ "notsig"),
    change = factor(change, levels = c("up", "down", "notsig"))
  ) %>% 
  lapply(arrange,
         desc(abs(logfc)))

timepoint.de.sig.list <- lapply(timepoint.de.list, filter, change != "notsig")
timepoint.de.sig.list
```

### DE Volcano plots

```{r, fig.width=8, fig.height=6}
timepoint.volcano.plots <- list()
de.list.plot.by <- timepoint.de.list
for (i in seq_along(de.list.plot.by)) {
  timepoint.volcano.plots[[names(de.list.plot.by)[[i]]]] <-
    ggplot(de.list.plot.by[[i]], aes(
      x = logfc,
      y = -log10(pvalue),
      color = change
    )) +
    geom_point() +
    scale_color_manual(values = c("red", "blue", "gray50")) +
    geom_hline(
      yintercept = -log10(sig.pvalue),
      linetype = "dashed",
      color = "firebrick"
    ) +
    geom_vline(
      xintercept = c(-sig.logfc, sig.logfc),
      linetype = "dashed",
      color = "firebrick"
    ) +
    geom_point(data = filter(de.list.plot.by[[i]], change != "notsig")) +
    
    ggrepel::geom_text_repel(
      data = filter(de.list.plot.by[[i]], change != "notsig"),
      aes(label = gene),
      show.legend = FALSE
    ) +
    
    labs(
      title = names(de.list.plot.by)[[i]],
      x = expression( ~ Log[2] ~ "(fold change)"),
      y = expression("-" ~ Log[10] ~ "(p value)")
    ) +
    scale_x_continuous(trans = "pseudo_log")
}
timepoint.volcano.plots
```

### Gene heatmaps for all patients

```{r, fig.width=5, fig.height=8}
patient.test = "P009"

genes.to.plot <- timepoint.de.list[[patient.test]] %>% 
  filter(change != "notsig") %>% 
  arrange(desc(abs(stat))) %>% 
  pull(gene)

sce.MCL[, sce.MCL$patient == patient.test] %>%
  plotGroupedHeatmap(
    .,
    features = genes.to.plot,
    group = "CNV.clone",
    center = TRUE,
    main = patient.test,
    fontsize_row = 8,
    zlim = c(-0.1, 0.1),
    cluster_rows = T
  )
```

```{r, fig.width=5, fig.height=8}
patient.test = "P022"

genes.to.plot <- timepoint.de.list[[patient.test]] %>% 
  filter(change != "notsig") %>% 
  arrange(desc(abs(stat))) %>% 
  pull(gene)

sce.MCL[, sce.MCL$patient == patient.test] %>%
  plotGroupedHeatmap(
    .,
    features = genes.to.plot,
    group = "CNV.clone",
    center = TRUE,
    main = patient.test,
    fontsize_row = 8,
    zlim = c(-0.5, 0.5),
    cluster_rows = T
  )
```

```{r, fig.width=5, fig.height=8}
patient.test = "P027"

genes.to.plot <- timepoint.de.list[[patient.test]] %>% 
  filter(change != "notsig") %>% 
  arrange(desc(abs(stat))) %>% 
  pull(gene)

sce.MCL[, sce.MCL$patient == patient.test] %>%
  plotGroupedHeatmap(
    .,
    features = genes.to.plot,
    group = "CNV.clone",
    center = TRUE,
    main = patient.test,
    fontsize_row = 8,
    zlim = c(-0.5, 0.5),
    cluster_rows = T
  )
```

```{r, fig.width=5, fig.height=8}
patient.test = "P087"

genes.to.plot <- timepoint.de.list[[patient.test]] %>% 
  filter(change != "notsig") %>% 
  arrange(desc(abs(stat))) %>% 
  pull(gene)

sce.MCL[, sce.MCL$patient == patient.test] %>%
  plotGroupedHeatmap(
    .,
    features = genes.to.plot,
    group = "CNV.clone",
    center = TRUE,
    main = patient.test,
    fontsize_row = 8,
    zlim = c(-0.5, 0.5),
    cluster_rows = T
  )
```

### DE overlap between patients

```{r}
timepoint.de.sig.overlap <- timepoint.de.sig.list %>% 
  bind_rows(.id = "patient") %>% 
  mutate(
    patient = factor(patient, levels = patients)
  ) %>% 
  dplyr::select(patient, gene, logfc) %>% 
  pivot_wider(names_from = patient, values_from = logfc) %>% 
  filter(rowSums(is.na(.)) == 1) 

timepoint.de.sig.overlap %>% 
  filter(
    (is.na(P009) | P009 > 0) &
      (is.na(P022) | P022 > 0) &
      (is.na(P027) | P027 > 0) &
      (is.na(P087) | P087 > 0)
  )
```

```{r}
timepoint.de.sig.overlap %>% 
  filter(
    (is.na(P009) | P009 < 0) &
      (is.na(P022) | P022 < 0) &
      (is.na(P027) | P027 < 0) &
      (is.na(P087) | P087 < 0)
  )
```

```{r}
timepoint.de.sig.list %>% 
  bind_rows(.id = "patient") %>% 
  mutate(
    patient = factor(patient, levels = patients)
  ) %>% 
  dplyr::select(patient, gene, logfc) %>% 
  pivot_wider(names_from = patient, values_from = logfc) %>% 
  filter(rowSums(is.na(.)) == 0)
```

```{r, fig.width=7, fig.height=4}
plotExpression(sce.MCL.plot,
               features = c("GAS6", "CCL3L1", "CMTM4", "NTNG1"),
               x = "timepoint", color_by = "timepoint"
               ) + labs(x = NULL)
```

```{r, fig.width=7, fig.height=6}
plotExpression(
  sce.MCL.plot,
  features = c("JAKMIP2", "RAB20", "CDHR3", "ZBTB16", "PELI2"),
  x = "timepoint",
  color_by = "timepoint"
)
```

### GSEA by `DECENT` results

#### KEGG

```{r}
DE.kegg.out.list <- list()
for (current_patient in samples) {
  selected.markers <- timepoint.de.list[[current_patient]]
  
  is.de <- which(selected.markers$change != "notsig")
  
  entrez.ids <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = selected.markers$gene,
    column = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  DE.kegg.out.list[[current_patient]] <-
    limma::kegga(unique(entrez.ids[is.de]),
                 species = "Hs",
                 universe = unique(entrez.ids)) %>%
    rownames_to_column("PathwayID") %>%
    filter(!str_detect(PathwayID, "hsa05")) %>%
    arrange(P.DE) %>%
    filter(P.DE < 0.01)
}
DE.kegg.out.list
```

```{r}
bind_rows(DE.kegg.out.list, .id = "patient") %>%
  dplyr::select(patient, Pathway, DE) %>%
  pivot_wider(names_from = patient, values_from = DE) %>%
  filter(rowSums(is.na(.)) < 1)
```

#### Reactome

```{r}
DECENT.GSE.list <- list()
for (current_patient in samples) {
  genes.table <- timepoint.de.list[[current_patient]] %>% 
    arrange(desc(stat))
  
  entrez.ids <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = genes.table$gene,
    column = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  gene.ranks <- genes.table$stat
  names(gene.ranks) <- entrez.ids
  gene.ranks <- gene.ranks[!is.na(entrez.ids)]
  
  DECENT.GSE.list[[current_patient]] <- gsePathway(
    gene.ranks[unique(names(gene.ranks))],
    organism = "human",
    pvalueCutoff = 0.01,
    pAdjustMethod = "BH",
    eps = 0,
    seed = TRUE
  )
}
lapply(DECENT.GSE.list, as.data.frame)
```

```{r, fig.width=9, fig.height=4}
DECENT.GSE.joined <- lapply(DECENT.GSE.list, as.data.frame) %>%
  bind_rows(., .id = "patient") %>%
  remove_rownames() %>%
  dplyr::select(patient, Description, NES) %>%
  pivot_wider(names_from = patient, values_from = NES) %>%
  filter(rowSums(is.na(.)) < 2)

DECENT.GSE.mat <- DECENT.GSE.joined %>%
  column_to_rownames("Description") %>%
  as.matrix()

DECENT.GSE.joined
```

# Comparison of pseudobulk DE and clonal DE

```{r message=FALSE}
DE.output.path <- file.path(wd, "scRNAseq/analysis/outputs/DE_DECENT_clone")

clone.de.list <- list()
for(samplename in samples){
  filename <- paste0("DE_clone_", samplename, ".csv")
  clone.de.list[[samplename]] <- read_csv(file.path(DE.output.path, samplename, filename))
}

clone.de.list <- lapply(
    clone.de.list,
    mutate,
    change = case_when(abs(logfc) > sig.logfc & pvalue < sig.pvalue & logfc < 0 ~ "down",
                       abs(logfc) > sig.logfc & pvalue < sig.pvalue & logfc > 0 ~ "up",
                       TRUE ~ "notsig"),
    change = factor(change, levels = c("up", "down", "notsig"))
  ) %>% 
  lapply(arrange,
         desc(abs(logfc)))

clone.de.sig.list <- lapply(clone.de.list, filter, change != "notsig")
clone.de.sig.list
```

```{r}
joined.de.sig.list <- list()
for(sample_name in samples){
  joined.de.sig.list[[sample_name]] <- full_join(
    clone.de.sig.list[[sample_name]] %>% dplyr::select(gene, change),
    timepoint.de.sig.list[[sample_name]] %>% dplyr::select(gene, change),
    by = "gene", suffix = c("_clone", "_timepoint")
  ) %>% 
    arrange(change_clone, change_timepoint, gene)
}
joined.de.sig.list
```

```{r}
library(VennDiagram)

Venn.output.path <- file.path(wd, "scRNAseq/analysis/results/DE_overlaps")
dir.create(Venn.output.path, showWarnings = FALSE)

for (i in seq_along(joined.de.sig.list)) {
  genes.clone <- joined.de.sig.list[[i]] %>%
    filter(!is.na(change_clone)) %>%
    pull(gene)
  genes.timepoint <- joined.de.sig.list[[i]] %>%
    filter(!is.na(change_timepoint)) %>%
    pull(gene)
  
  filename <-
    paste0("DE_intersect_", names(joined.de.sig.list)[i], ".png")
  venn.diagram(
    main = names(joined.de.sig.list)[i],
    main.fontface = "bold",
    main.fontfamily = "sans",
    
    x = list(genes.timepoint, genes.clone),
    category.names = c("Timepoint", "Clone"),
    fill = c("blue", "red"),
    filename = file.path(Venn.output.path, filename),
    height = 1500, width = 3000, units = "px",
    
    cat.fontfamily = "sans",
    cat.pos = c(-20, 20)
  )
}
```






