---
title: "DECENT DE analysis"
output: html_notebook
---

```{r setup}
library(tidyverse)
library(openxlsx)
library(scater)
library(scran)
library(ReactomePA)
library(ComplexHeatmap)

# Path to project folder
wd <- here::here()
source(file.path(wd, "scRNAseq/analysis/scripts/colors.R"))
source(file.path(wd, "scRNAseq/analysis/scripts/setup.R"))

# Parallelization
library(BiocParallel)
bpp <- MulticoreParam()
```

```{r}
iteration_input = "6_MCL_CNV"
sce.dir <- file.path(wd, "scRNAseq/analysis/outputs", iteration_input)
sce.MCL <- readRDS(file = file.path(sce.dir, "sce.rds"))
sce.MCL
```

```{r, fig.width=12, fig.height=6}
sce.MCL.plot <- sce.MCL
sce.MCL.plot$Sample <- sce.MCL.plot$sample_norep_plot
sce.MCL.plot$Sample <- factor(sce.MCL.plot$Sample, 
                              levels = levels(sce.MCL.plot$sample_norep) %>% str_replace_all(., "_", " "))

plotReducedDim(sce.MCL.plot,
               dimred = "TSNE_corrected_SamplePhase",
               color_by = "patient",
               other_fields = "timepoint") +
  facet_wrap( ~ timepoint) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  xlim(-35, 40) + ylim(-40, 35) +
  cowplot::theme_half_open(font_size = 14) +
  theme(strip.text = element_text(face = "bold")) +
  cowplot::panel_border(color = "grey50") +
  labs(x = "TSNE 1", y = "TSNE 2", color = "Patient")
```

```{r, fig.width=11, fig.height=5.5, message=FALSE, warning=FALSE}
plotReducedDim(sce.MCL.plot,
               dimred = "TSNE_corrected_SamplePhase",
               color_by = "Sample",
               other_fields = "timepoint") +
  facet_wrap( ~ timepoint) +
  theme_bw() +
  scale_color_manual(values = get_palettes("tableau20")) +
  labs(color = "Sample") +
  xlim(-35, 40) + ylim(-40, 35) +
  guides(color = guide_legend(override.aes = list(size = 4))) +
  cowplot::theme_half_open(font_size = 16) +
  cowplot::panel_border(color = "grey50") +
  theme(strip.text = element_text(face = "bold"))
```

# Pseudobulk DE between clones

## Output of `DECENT` package

This output was pre-calculated. `DECENT` DE analysis outputs a table of genes compared between conditions with log fold changes, p values and stat value. Here that table is annotated and then plotted.

```{r message=FALSE}
DE.output.path <- file.path(wd, "scRNAseq/analysis/outputs/DE_DECENT_clone")

clone.de.list <- list()
samples <- patients
for(samplename in samples){
  filename <- paste0("DE_clone_", samplename, ".csv")
  clone.de.list[[samplename]] <- read_csv(file.path(DE.output.path, samplename, filename))
}

sig.pvalue = 0.01
sig.logfc = 2

clone.de.list <- lapply(
    clone.de.list,
    mutate,
    change = case_when(abs(logfc) > sig.logfc & pvalue < sig.pvalue & logfc < 0 ~ "down",
                       abs(logfc) > sig.logfc & pvalue < sig.pvalue & logfc > 0 ~ "up",
                       TRUE ~ "notsig"),
    change = factor(change, levels = c("up", "down", "notsig"))
  ) %>% 
  lapply(arrange,
         desc(abs(logfc)))

clone.de.sig.list <- lapply(clone.de.list, filter, change != "notsig")
clone.de.sig.list
```

```{r, fig.width=8, fig.height=6}
timepoint.volcano.plots <- list()
de.list.plot.by <- clone.de.list
for (i in seq_along(de.list.plot.by)) {
  timepoint.volcano.plots[[names(de.list.plot.by)[[i]]]] <-
    ggplot(de.list.plot.by[[i]], aes(
      x = logfc,
      y = -log10(pvalue),
      color = change
    )) +
    geom_point() +
    scale_color_manual(values = c("red", "blue", "gray50")) +
    geom_hline(
      yintercept = -log10(sig.pvalue),
      linetype = "dashed",
      color = "firebrick"
    ) +
    geom_vline(
      xintercept = c(-sig.logfc, sig.logfc),
      linetype = "dashed",
      color = "firebrick"
    ) +
    geom_point(data = filter(de.list.plot.by[[i]], change != "notsig")) +
    
    ggrepel::geom_text_repel(
      data = filter(de.list.plot.by[[i]], change != "notsig"),
      aes(label = gene),
      show.legend = FALSE
    ) +
    
    labs(
      title = names(de.list.plot.by)[[i]],
      x = expression( ~ Log[2] ~ "(fold change)"),
      y = expression("-" ~ Log[10] ~ "(p value)")
    ) +
    scale_x_continuous(trans = "pseudo_log")
}
timepoint.volcano.plots
```

### Check a specific gene

```{r, fig.width=6, fig.height=4}
patient.test = "P087"
gene.test <- "BCL2L10"

plotExpression(
  sce.MCL[, sce.MCL$patient == patient.test],
  features = gene.test,
  x = "CNV.group", color_by = "CNV.clone",
  other_fields = "timepoint", ncol = 4
) +
  facet_wrap(~ timepoint, ncol = 1) +
  labs(title = gene.test)
```

```{r, fig.width=12, fig.height=4}
plotReducedDim(sce.MCL[, sce.MCL$patient == patient.test],
               dimred = "TSNE_corrected_SamplePhase",
               color_by = gene.test,
               other_fields = "CNV.clone") +
  facet_wrap(~ CNV.clone)
```

### Gene heatmaps for all patients

```{r, fig.width=5, fig.height=8}
ht.list <- list()
for(patient.test in patients){
  genes.to.plot <- clone.de.list[[patient.test]] %>% 
  filter(change != "notsig") %>% 
  arrange(desc(abs(stat))) %>% 
  pull(gene)
  
  ht.list[[patient.test]] <- sce.MCL[, sce.MCL$patient == patient.test] %>%
  plotGroupedHeatmap(
    .,
    features = genes.to.plot,
    group = "CNV.clone",
    center = TRUE,
    main = patient.test,
    fontsize_row = 8,
    zlim = c(-0.1, 0.1),
    cluster_rows = T
  )
  
  print(
    grid::grid.draw(ht.list[[patient.test]]$gtable)
  )
}
```

### DE overlap between patients

```{r}
clone.de.sig.list %>% 
  bind_rows(.id = "patient") %>% 
  mutate(
    patient = factor(patient, levels = patients)
  ) %>% 
  dplyr::select(patient, gene, logfc) %>% 
  pivot_wider(names_from = patient, values_from = logfc) %>% 
  filter(rowSums(is.na(.)) < 3)
```

```{r, fig.width=10, fig.height=12}
plotReducedDim(
  sce.MCL,
  dimred = "TSNE_corrected_SamplePhase",
  color_by = "PPM1L",
  other_fields = c("patient", "CNV.clone")
) +
  facet_grid(patient ~ CNV.clone) +
  theme_bw()
```

## GSEA of `DECENT` DE results

### KEGG

```{r}
DE.kegg.out.list <- list()
for (current_patient in patients) {
  selected.markers <- clone.de.list[[current_patient]]
  
  is.de <- which(selected.markers$change != "notsig")
  
  entrez.ids <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = selected.markers$gene,
    column = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  DE.kegg.out.list[[current_patient]] <-
    limma::kegga(unique(entrez.ids[is.de]),
                 species = "Hs",
                 universe = unique(entrez.ids)) %>%
    rownames_to_column("PathwayID") %>%
    filter(!str_detect(PathwayID, "hsa05")) %>%
    arrange(P.DE) %>%
    filter(P.DE < 0.01)
}
DE.kegg.out.list
```

### Reactome

```{r}
DECENT.GSE.list <- list()
for (current_patient in patients) {
  genes.table <- clone.de.list[[current_patient]] %>% 
    arrange(desc(stat))
  
  entrez.ids <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = genes.table$gene,
    column = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  gene.ranks <- genes.table$stat
  names(gene.ranks) <- entrez.ids
  gene.ranks <- gene.ranks[!is.na(entrez.ids)]
  
  DECENT.GSE.list[[current_patient]] <- gsePathway(
    gene.ranks[unique(names(gene.ranks))],
    organism = "human",
    pvalueCutoff = 0.01,
    pAdjustMethod = "BH",
    eps = 0,
    seed = TRUE
  )
}

# save as object for plotting
saveRDS(DECENT.GSE.list, file = file.path(wd, "scRNAseq/analysis/results",
                                          "DECENT.GSE.RPA.rds"))

lapply(DECENT.GSE.list, as.data.frame)
```

```{r}
lapply(DECENT.GSE.list[1:4], as.data.frame) %>% 
  lapply(., dplyr::select, Description, NES, p.adjust) %>% 
  lapply(., remove_rownames) %>% 
  lapply(., arrange, desc(NES))
```

#### GSEA overlap heatmap

```{r, fig.width=9, fig.height=4}
DECENT.GSE.joined <- lapply(DECENT.GSE.list, as.data.frame) %>%
  bind_rows(., .id = "patient") %>%
  remove_rownames() %>%
  dplyr::select(patient, Description, NES) %>%
  pivot_wider(names_from = patient, values_from = NES) %>%
  filter(rowSums(is.na(.)) < 2)

DECENT.GSE.mat <- DECENT.GSE.joined %>%
  column_to_rownames("Description") %>%
  as.matrix()

ht <- Heatmap(
  matrix = DECENT.GSE.mat,
  name = "NES",
  na_col = "gray70",
  cluster_rows = F,
  cluster_columns = F,
  row_names_side = "left",
  column_names_side = "top",
  column_names_rot = 0,
  column_names_centered = T,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.1f", DECENT.GSE.mat[i, j]), x, y, gp = gpar(fontsize = 10))
  }
)

draw(ht, padding = unit(c(2, 95, 2, 2), "mm"))
```

## Marker gene detection by `scran` (per OSCA)

```{r}
list.markers.DGs <- list()
for (current_patient in patients) {
  sce.test <- sce.MCL[, sce.MCL$patient == current_patient]
  sce.test <- sce.test[, sce.test$timepoint == "DG"]
  sce.test$CNV.clone <-
    factor(sce.test$CNV.clone, levels = c("DGex", "DGs"))
  
  markers <- scoreMarkers(sce.test,
                          groups = sce.test$CNV.clone,
                          BPPARAM = bpp)
  
  list.markers.DGs[[current_patient]] <- markers[["DGs"]] %>%
    as.data.frame() %>%
    arrange(desc(abs(mean.logFC.detected))) %>%
    rownames_to_column("gene")
}
list.markers.DGs
```

```{r}
mean.logFC.threshold <- 2

list.markers.DGs.sig <- list.markers.DGs %>% 
  lapply(., function (x) filter(x, abs(mean.logFC.detected) > mean.logFC.threshold)) %>% 
  lapply(., arrange, rank.AUC)
list.markers.DGs.sig
```

```{r, fig.width=6.5, fig.height=5}
list.markers.plots <- list()
for(current_patient in patients){
  list.markers.plots[[current_patient]] <- list.markers.DGs[[current_patient]] %>% 
    ggplot(aes(x = mean.logFC.cohen, y = mean.AUC)) +
    geom_point(color = "gray20") +
    geom_point(
      data = list.markers.DGs.sig[[current_patient]],
      aes(color = mean.logFC.detected)) +
    scale_color_gradient2(low = "blue", high = "red") +
    theme_bw() +
    labs(title = current_patient) +
    ylim(0, 1) + xlim(-3, 3)
}
list.markers.plots
```

```{r, fig.width=5, fig.height=15}
list.marker.heatmaps <- list()
for(current_patient in patients){
list.marker.heatmaps[[current_patient]] <- sce.MCL[, sce.MCL$patient == current_patient] %>% 
  plotGroupedHeatmap(.,
                     features = list.markers.DGs.sig[[current_patient]]$gene,
                     group="CNV.clone",
                     main = current_patient,
                     center=TRUE,
                     fontsize_row = 6
                     ) %>% ggplotify::as.ggplot() +
  labs(title = current_patient)
}
```

This looks similar to DECENT but worse.

### GSEA by `scran` markers

#### KEGG

```{r}
scran.kegg.out.list <- list()

for (current_patient in patients) {
  selected.markers <- list.markers.DGs[[current_patient]]
  
  entrez.ids <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = selected.markers$gene,
    column = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  is.de <- which(selected.markers$gene %in% list.markers.DGs.sig[[current_patient]]$gene)
  
  scran.kegg.out.list[[current_patient]] <-
    limma::kegga(unique(entrez.ids[is.de]),
                 species = "Hs",
                 universe = unique(entrez.ids)) %>%
    rownames_to_column("PathwayID") %>%
    filter(!str_detect(PathwayID, "hsa05")) %>%
    arrange(P.DE) %>%
    filter(P.DE < 0.01)
}
scran.kegg.out.list
```

#### Reactome

```{r}
scran.PEA.list <- list()
for (current_patient in patients) {
  selected.markers <- list.markers.DGs[[current_patient]] %>% 
    arrange(rank.logFC.detected)
  
  entrez.ids <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = selected.markers$gene,
    column = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  scran.PEA.list[[current_patient]] <- enrichPathway(
                       gene = entrez.ids,
                       organism = "human",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH",
                       qvalueCutoff = 0.2
                     )
}
lapply(scran.PEA.list, as.data.frame)
```

```{r}
scran.GSE.list <- list()
for (current_patient in patients) {
  genes.table <- list.markers.DGs[[current_patient]] %>%
    arrange(rank.logFC.detected) %>%
    dplyr::select(gene, mean.logFC.detected, rank.logFC.detected)
  
  entrez.ids <- AnnotationDbi::mapIds(
    org.Hs.eg.db::org.Hs.eg.db,
    keys = genes.table$gene,
    column = "ENTREZID",
    keytype = "SYMBOL"
  )
  
  gene.ranks <- genes.table$mean.logFC.detected
  names(gene.ranks) <- entrez.ids
  gene.ranks <- gene.ranks[!is.na(entrez.ids)]
  
  scran.GSE.list[[current_patient]] <- gsePathway(
    gene.ranks,
    organism = "human",
    pvalueCutoff = 0.01,
    pAdjustMethod = "BH",
    nPermSimple = 10000,
    eps = 0
  )
}
lapply(scran.GSE.list, as.data.frame)
```

```{r, fig.width=6, fig.height=8}
lapply(scran.GSE.list, dotplot, showCategory = 15)
```

# Comparison of `DECENT` and `scran` DE gene detection

```{r}
joined.sig.de <- list()
for(current_patient in patients){
  joined.sig.de[[current_patient]] <- full_join(
    dplyr::select(clone.de.sig.list[[current_patient]], gene, "DECENT.lfc" = logfc),
    dplyr::select(list.markers.DGs.sig[[current_patient]], gene, "markers.lfc" = mean.logFC.detected),
    by = join_by(gene)
  )
}
joined.sig.de
```

```{r}
joined.sig.de.mat <- lapply(joined.sig.de, na.omit) %>% 
  bind_rows(.id = "patient") %>% 
  dplyr::select(-DECENT.lfc) %>% 
  mutate(
    patient = factor(patient, levels = patients)
  ) %>% 
  pivot_wider(names_from = patient, values_from = markers.lfc) %>% 
  filter(rowSums(is.na(.)) < 4)
joined.sig.de.mat
```


